---
title: "01_GO"
documentclass: ctexart
output:
  bookdown::pdf_book:
    latex_engine: xelatex
  html_document: default
date: "2024-05-28"
---

# GO enrichment analysis {#clusterprofiler-go}

GO comprises three orthogonal ontologies, i.e. molecular function (MF), biological process (BP), and cellular component (CC).

GO（Gene Ontology，基因本体论）包括三个正交的本体（指的是在基因本体论中三个不同的层面或方面），即分子功能（MF，Molecular Function）、生物学过程（BP，Biological Process）和细胞组分（CC，Cellular Component）。 这些本体是相互独立的，每一个都描述了基因和蛋白质的不同方面。 "Orthogonal" 的意思是"正交"，在这里指的是这三个本体是相互独立的，彼此之间没有重叠或依赖关系。换句话说，一个基因或蛋白质可以在每个本体中被描述，而不会因为在一个本体中的描述而影响到在另一个本体中的描述。

```{r include=FALSE}
library(knitr)
opts_chunk$set(message=FALSE, warning=FALSE, eval=TRUE, echo=TRUE, cache=TRUE)
library(clusterProfiler)
```

## Supported organisms {#clusterProfiler-go-supported-organisms}

GO analyses (`groupGO()`, `enrichGO()` and `gseGO()`) support organisms that have an `OrgDb` object available (see also [session 2.2](#gosemsim-supported-organisms)).

If a user has GO annotation data (in a `data.frame` format with the first column as gene ID and the second column as GO ID), they can use the `enricher()` and `GSEA()` functions to perform an over-representation test and gene set enrichment analysis.

If the genes are annotated by direction annotation, they should also be annotated by their ancestor GO nodes (indirect annotation). If a user only has direct annotation, they can pass their annotation to the `buildGOmap` function, which will infer indirect annotation and generate a `data.frame` that is suitable for both `enricher()` and `GSEA()`.

## GO classification

In `clusterProfiler`, the `groupGO()` function is designed for gene classification based on GO distribution at a specific level. Here we use the dataset `geneList` provided by `"DOSE"`.

```{r clusterprofiler-groupgo, warning=FALSE}
library(clusterProfiler)
deg = read.table("outputs/deg_all.txt")

df <- readxl::read_xlsx('outputs/MAIT激活方式marker.xlsx')
df$gene <- tolower(df$IFNG)
df$gene <- paste0(toupper(substring(df$gene, 1, 1)), substring(df$gene, 2))
df$IFNG <- NULL
df$logFC <- deg[df$gene, "logFC"]
df <- na.omit(df)
geneList <- df$logFC
names(geneList) <- df$gene
geneList <- sort(geneList, decreasing = T)
saveRDS(geneList, file = "outputs/geneList.rds")

# gene <- names(geneList)[abs(geneList) > 2]
gene <- names(geneList)

# Entrez gene ID
head(gene)

organisms = "org.Rn.eg.db"
library(organisms, character.only = T)

ggo <- groupGO(gene     = gene,
               OrgDb    = organisms,
               keyType = "SYMBOL",
               ont      = "CC",
               level    = 3,
               readable = TRUE)

head(ggo)
```

The `gene` parameter is a vector of gene IDs (can be any ID type that is supported by the corresponding `OrgDb`, see also [session 16.1](#id-convert)). If `readable` is set to `TRUE`, the input gene IDs will be converted to gene symbols.

## GO over-representation analysis {#clusterprofiler-go-ora}

The `"clusterProfiler"` package implements `enrichGO()` for gene ontology over-representation test.

```{r}
ego <- enrichGO(gene          = gene,
                # universe      = names(geneList),
                OrgDb         = organisms,
                keyType = "SYMBOL",
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = T)
head(ego)
```

Any gene ID type that is supported in `OrgDb` can be directly used in GO analyses. Users need to specify the `keyType` parameter to specify the input gene ID type.

```{r}
gene.df <- bitr(gene, fromType = "SYMBOL",
                toType = c("ENSEMBL", "SYMBOL"),
                OrgDb = organisms)

ego2 <- enrichGO(gene         = gene.df$ENSEMBL,
                OrgDb         = organisms,
                keyType       = 'ENSEMBL',
                ont           = "CC",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05)
head(ego2, 3)
```

Gene IDs can be mapped to gene Symbols by using the parameter `readable=TRUE` or [`setReadable()` function](#setReadable).

## GO Gene Set Enrichment Analysis

The `"clusterProfiler"` package provides the `gseGO()` function for [gene set enrichment analysis](#gsea-algorithm) using gene ontology.

```{r eval=FALSE}
ego3 <- gseGO(geneList     = geneList,
              OrgDb        = organisms,
              keyType = "SYMBOL",
              ont          = "CC",
              minGSSize    = 10,
              maxGSSize    = 500,
              pvalueCutoff = 0.05,
              verbose      = FALSE)
head(ego3)
```

The format of input data, `geneList`, was documented in the [FAQ](#genelist). Beware that only gene Set size in `[minGSSize, maxGSSize]` will be tested.

## GO analysis for non-model organisms {#clusterprofiler-go-non-model}

Both the `enrichGO()` and `gseGO()` functions require an `OrgDb` object as the background annotation. For organisms that don't have `OrgDb` provided by `Bioconductor`, users can query one (if available) online via `"AnnotationHub"`. If there is no `OrgDb` available, users can obtain GO annotation from other sources, e.g. from `"biomaRt"`, or annotate the genes using [Blast2GO](https://www.blast2go.com/) or the [Trinotate](https://rnabio.org/module-07-trinotate/0007/02/01/Trinotate/) pipeline. Then the `enricher()` or `GSEA()` functions can be used to perform GO analysis for these organisms, similar to the examples using wikiPathways and MSigDB. Another solution is to create an `OrgDb` on your own using `"AnnotationForge"` package.

Here is an example of querying GO annotation from Ensembl using `"biomaRt"`.

```{r eval=FALSE}
library(biomaRt)
ensembl <- useEnsemblGenomes(biomart = "plants_mart", dataset = "nattenuata_eg_gene")
gene2go <- getBM(attributes =c("ensembl_gene_id", "go_id"), mart=ensembl)
```

## Visualize enriched GO terms as a directed acyclic graph

The `goplot()` function can accept the output of `enrichGO` and visualize the enriched GO induced graph.

(ref:goplotscap) Goplot of enrichment analysis.

(ref:goplotcap) **Goplot of enrichment analysis.**

```{r goplot, fig.height=12, fig.width=8, fig.cap="(ref:goplotcap)", fig.scap="(ref:goplotscap)"}
goplot(ego)
```

## Summary {#clusterprofiler-go-summary}

GO semantic similarity can be calculated by `"GOSemSim"` [@yu2010]. We can use it to cluster genes/proteins into different clusters based on their functional similarity and can also use it to measure the similarities among GO terms to reduce the redundancy of GO enrichment results.
